Concurrent Programming in Harmony

### 2020-09-30 book/paper.tex
Latest
Fixed Hoare monitor semantics

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inductive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
For simplicity, we will suppose as a given that 
$\mathtt{process}(i)@cs \Rightarrow \mathit{flags}[i]$, the truth of which is
easy to see from the code.
Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
\mathit{turn} = i$, that is, the condition on the \textbf{await}
statement for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following:
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ holds or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}%
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \textit{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
\index{condition variable}%
There are two operations on condition variables: \texttt{wait}
\index{wait}%
and
\texttt{signal}.
\index{signal}%
If $c$ is a pointer to a condition variable, then the semantics of the
operations, in Harmony, are as follows:

\begin{code}
\harmonysource{hoaremon}
\end{code}

\texttt{wait} releases the mutex and blocks while trying
to procure the condition variable (which is a semaphore that is 0 when
\texttt{wait} is invoked).
\texttt{signal} vacates the condition variable, passing the baton to
a process trying to procure it, and then procures the mutex.
The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
\texttt{V} operations alternate as prescribed by the SBS technique.
Our implementations of reader/writer locks can be easily changed to
use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
and \texttt{V}.

>>>>

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inductive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
\mathit{turn} = i$, that is, the condition on the \textbf{await}
statement for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following:
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ holds or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

\begin{figure}
\begin{code}
\harmonysource{hoare}
\end{code}
\caption{\harmonylink{code/hoare.hny} Implementation of Hoare monitors.}
\label{fig:hoare}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{hoaretest}
\end{code}
\caption{\harmonylink{code/hoaretest.hny} Example of Hoare monitors.}
\label{fig:hoaretest}
\end{figure}

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}%
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \textit{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are encapsulated in \emph{condition variables}.
\index{condition variable}%
Each condition variable also keeps track of the number of processes waiting
on the condition variable.
There are two operations on condition variables: \texttt{wait}
\index{wait}%
and
\texttt{signal}.
\index{signal}%

\autoref{fig:hoare} presents the semantics by giving an implementation of
\texttt{wait} and \texttt{signal} as well as the code that is used to enter
and exit monitor methods.
\autoref{fig:hoaretest} presents a simple example lifted directly from
the original paper~\cite{Hoare74}.
In the design proposed by Hoare, the monitor methods automatically acquire
and release the monitor mutex.

\texttt{wait} increments the condition's counter, releases the monitor mutex,
blocks while trying to procure the condition's semaphore, and finally decrements
the counter---in much the same way as we have seen for SBS.
\texttt{signal} checks to see if the condition's count is non-zero, and if so
vacates the condition's semaphore---passing the baton to
a process trying to procure it---and then blocks by procuring the mutex again.
The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
\texttt{V} operations alternate as prescribed by the SBS technique.
Our implementations of reader/writer locks can be easily changed to
use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
and \texttt{V}.
